Exploiting Array Syntax in Fortran for Accelerator Programming

### 2010-12-20 opencl/papers/intro.tex
updates
tweaks to shrink to 8 pages

%We provide (via Fortran interfaces in the ForOpenCL library) a
%mechanism to call the C OpenCL runtime and enable Fortran programmers
%to access OpenCL kernels.  
Transformations are supplied that provide a mechanism for converting
Fortran procedures written in the Fortran subset described in this
paper to OpenCL kernels.  We use the ROSE compiler infrastructure
\cite{rose-web} to develop these transformations.  ROSE uses the Open
Fortran Parser \cite{ofp:sf} to parse Fortran 2008 syntax and can
generate C-based OpenCL.  Since ROSE's intermediate representation
(IR) was constructed to represent multiple languages, it is relatively
straightforward to transform high-level Fortran IR nodes to C OpenCL
nodes.

@@

%% NEW - CER
History has also shown that an investment in rewriting code does not
guarantee success either, as seen in an effort at LANL to modernize a
legacy Fortran code with the newer C++ POOMA framework.  This
massive overhaul effort led to a code that was both slower and less
flexible than the original Fortran \cite{basili08hpc}.  Similar
experiences have occurred in the past, notably during the development
of the functional SISAL language in the early 1990s.

@@

Similar whole array shifting was used in higher dimensions for finite
different codes within the computational physics community, especially
at Los Alamos for codes targeting the CM-5 system that resided there until the
late 1990s.

>>>>

%We provide (via Fortran interfaces in the ForOpenCL library) a
%mechanism to call the C OpenCL runtime and enable Fortran programmers
%to access OpenCL kernels.  
Transformations are supplied that provide a mechanism for converting
Fortran procedures written in the Fortran subset described in this
paper to OpenCL kernels.  We use the ROSE compiler
infrastructure\footnote{\url{http://www.rosecompiler.org/}} to
develop these transformations.  ROSE uses the Open Fortran
Parser\footnote{\url{http://fortran-parser.sf.net/}} to parse
Fortran 2008 syntax and can generate C-based OpenCL.  Since ROSE's
intermediate representation (IR) was constructed to represent multiple
languages, it is relatively straightforward to transform high-level
Fortran IR nodes to C OpenCL nodes.

@@

%% NEW - CER
History has also shown that an investment in rewriting code does not
guarantee success either, as seen in an effort at LANL to modernize a
legacy Fortran code with the newer C++ POOMA framework.  This
massive overhaul effort led to a code that was both slower and less
flexible than the original Fortran \cite{basili08hpc}.

%Similar
%experiences have occurred in the past, notably during the development
%of the functional SISAL language in the early 1990s.

@@

Similar whole array shifting was used in higher dimensions for finite
different codes within the computational physics community, especially
at Los Alamos for codes targeting the CM-5 system that resided there until the
late 1990s.  A body of research in compilation of stencil-based codes
that use shift operators targeting these systems is related to the
work we present here~\cite{stencil-compiler}.

### 2010-12-20 opencl/papers/source-to-source.tex
edits
tweaks to shrink to 8 pages

OpenCL~\cite{opencl08} is an open language standard for developing
applications for accelerators.  The C-based language provides extensions for
programming kernels that run on accelerator processing elements.  The kernels
are run by calling a C runtime library from the OpenCL host (normally the
CPU).  Efforts to standardize a C++ runtime are underway and Fortran
interfaces to the C runtime are described later.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{cl-memory.pdf}
\caption{A schematic of global memory for an array and its copy stored in local memory
for four thread groups.}
\label{fig:cl-memory}
\end{figure}

@@

\subsection{ForOpenCL}

This work describes transformations that automatically create OpenCL kernel
functions from Fortran pure and elemental procedures.  At present, these transformations
will not transform an entire program.  Users, for now, must explicitly replace
calls to Fortran kernel procedures that run on the device, with calls to the
OpenCL runtime that will run the kernel on the OpenCL host.  While these host
transformations are straightforward using ROSE, they are outside the scope of
this paper.

In addition to the Fortran to OpenCL transformations, the
ForOpenCL~\cite{ofp:sf} library provides programmers with the ability to
call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
providing Fortran 2003 interface descriptions and classes that allow language
interoperability with the OpenCL runtime.

@@

\subsubsection{interior and region functions}

@@

\subsubsection{function and variable declarations}

@@

\subsubsection{array syntax}

Array syntax transforms nearly directly to OpenCL code.  For example, interior
variables are particularly straightforward as they as scalar quantities in
OpenCL,

@@

{\small
\begin{verbatim}
Hx[l] = 0.5 * (region(H_local, face_lt)+ ...);
\end{verbatim}
}

\noindent
where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get\_local\_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get\_local\_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
OpenCL language standard.

>>>>

OpenCL~\cite{opencl08} is an open language standard for developing
applications for accelerators.
The kernels
are run by calling a C runtime library from the OpenCL host (normally the
CPU).  Efforts to standardize a C++ runtime are underway and Fortran
interfaces to the C runtime are described later.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=1.75in]{cl-memory.pdf}
\caption{A schematic of global memory for an array and its copy stored in local memory
for four thread groups.}
\label{fig:cl-memory}
\end{figure}

@@

%\subsection{ForOpenCL}

%This work describes transformations that automatically create OpenCL kernel
%functions from Fortran pure and elemental procedures.  At present, these transformations
%will not transform an entire program.  Users, for now, must explicitly replace
%calls to Fortran kernel procedures that run on the device, with calls to the
%OpenCL runtime that will run the kernel on the OpenCL host.  While these host
%transformations are straightforward using ROSE, they are outside the scope of
%this paper.

%In addition to the Fortran to OpenCL transformations, the
%ForOpenCL library available from the Open Fortran Parser web page provides programmers with the ability to
%call the C OpenCL runtime from Fortran.  ForOpenCL is a set of Fortran modules
%providing Fortran 2003 interface descriptions and classes that allow language
%interoperability with the OpenCL runtime.

@@

\subsubsection{Interior and region functions}

@@

\subsubsection{Function and variable declarations}

@@

\subsubsection{Array syntax}

Array syntax transforms nearly directly to OpenCL code.  For example, interior
variables are particularly straightforward as they are scalar quantities in
OpenCL,

@@

{\small
\begin{verbatim}
Hx[i] = 0.5 * (region(H_local, face_lt)+ ...);
\end{verbatim}
}

\noindent
where {\tt i = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get\_local\_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get\_local\_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
OpenCL language standard.
